Evolution of Spectral Distributions in Deep-Water Constant Vorticity Flows

### 2019-12-19 Writeup/higher_order_dispersion_relationship_nost_writeup.tex

To better understand why we must include the higher-order corrections to the dispersion used in the VDE, we examine how such terms are found from first principles.  To leading order from the kinetic equation over an infinitely deep fluid we have that 
\[
\eta_{t} = -\mathcal{H}Q,
\]
where $q(x,t) = \phi(x,\eta(x,t),t)$, $Q = q_{x}$, and $\mathcal{H}$ is the Hilbert transform.  Likewise, from the Bernoulli equation we have to leading order that
\[
Q _{t}+ \omega\eta_{t} + \eta_{x}= 0, 
\]
so that by combining the two expressions we have the wave equation in $\eta$ alone given by 
\begin{equation}
\eta_{tt}- \omega\mathcal{H}\eta_{t} - \mathcal{H}\eta_{x}= 0.
\label{linearwaveeq}
\end{equation}
Noting that 
\[
\left(-\mathcal{H} \pd_{x}\right)^{\widehat{}} = \left|k\right|,
\]
we then define the self-adjoint pseudo-differential operator $\mathcal{L}_{+}$ with symbol 
\[
\mathcal{\widehat{L}}_{+} = \sqrt{\left|k\right| }.
\]
Multiplying by $\eta_{t}$ and integrating over space gives us the conserved quantity
\[
\int_{\mathbb{R}} \left(\frac{1}{2}\eta_{t}^{2} + \frac{1}{2}\left(\mathcal{L}_{+}\eta \right)^{2}\right) dx = \tilde{E}.
\]
Note, 
\[
\int_{\mathbb{R}}\eta_{t}\mathcal{H}\eta_{t} dx = \frac{1}{2\pi}\int_{\mathbb{R}} i\mbox{sgn}(k) \left|\hat{\eta}_{t}(k,t) \right|^{2} dk = 0,
\]
since we assume $\eta_{t}$ is real.

Letting
\[
\eta(x,t) = \eta_{1}(\xi,\tau)e^{i\theta}+\mbox{cc}, ~ \xi = \epsilon(x+c_{g}t), ~ \tau = \epsilon^{2}t, ~ \theta = k_{0}x + \Omega(\omega,k_{0})t,
\]
and using the corresponding multiple-scales expansions
\[
\pd_{x} = ik_{0} + \epsilon \pd_{\xi}, ~ \pd_{t} = i\Omega + \epsilon c_{g}\pd_{\xi} + \epsilon^{2}\pd_{\tau}, 
\]
Equation \eqref{linearwaveeq} then becomes 
\begin{multline*}
\left(\Omega^{2} -s\omega\Omega - |k_{0}|\right)\eta_{1} + 2i\epsilon\left((2\Omega-s\omega)c_{g} - s\right)\pd_{\xi}\eta_{1}\\
+ \epsilon^{2}\left(i(2\Omega-s\omega)\pd_{\tau}+c^{2}_{g}\pd^{2}_{\xi} + 2\epsilon c_{g}\pd^{2}_{\xi \tau} + \epsilon^{2}\pd^{2}_{\tau}\right)\eta_{1} = 0.
\end{multline*}
Standard NLS theory has the first two terms vanishing, thus leaving us with the slowly evolving wave equation 
\[
\left(i(2\Omega-s\omega)\pd_{\tau}+c^{2}_{g}\pd^{2}_{\xi} + 2\epsilon c_{g}\pd^{2}_{\xi \tau} + \epsilon^{2}\pd^{2}_{\tau}\right)\eta_{1} = 0.
\]
Taking a Fourier transform in space so that $\pd_{\xi}\rightarrow i\tilde{k}$, we see the above wave equation propagates information temporally at the characteristic frequencies $\lambda_{\pm}$ where
\[
\lambda_{\pm} = \frac{(2\Omega-s\omega)}{2\epsilon^{2}}\left(-\left(1+\frac{2\epsilon c_{g}\tilde{k}}{2\Omega-s\omega}\right) \pm \left(1 + \frac{4\epsilon c_{g}\tilde{k}}{2\Omega-s\omega} \right)^{1/2} \right)
\]
Likewise, proceeding as above, we can readily derive the corresponding conserved quantity
\[
\frac{s\epsilon^{2}}{\omega-2s\Omega}\int_{\mathbb{R}}\left| \pd_{\tau}\eta_{1}\right|^{2}d\xi + \alpha_{d}\int_{\mathbb{R}} \left| \pd_{\xi}\eta_{1}\right|^{2}d\xi + \frac{\epsilon\tilde{\sigma}i}{\omega-2s\Omega}\int_{\mathbb{R}}\pd_{\xi}\eta^{\ast}_{1}\pd^{2}_{\xi}\eta_{1}d\xi= \bar{E}.
\]

\end{document}

>>>>

To better understand why we must include the higher-order corrections to the dispersion used in the VDE, we examine how such terms are found from first principles.  To leading order from the kinetic equation over an infinitely deep fluid we have that 
\[
\eta_{t} = -\mathcal{H}Q,
\]
where $q(x,t) = \phi(x,\eta(x,t),t)$, $Q = q_{x}$, and $\mathcal{H}$ is the Hilbert transform.  Likewise, from the Bernoulli equation we have to leading order that
\[
Q _{t}+ \omega\eta_{t} + \eta_{x}= 0, 
\]
so that by combining the two expressions we have the wave equation in $\eta$ alone given by 
\begin{equation}
\eta_{tt}- \omega\mathcal{H}\eta_{t} - \mathcal{H}\eta_{x}= 0.
\label{linearwaveeq}
\end{equation}
Using Fourier transforms, we can readily write a solution to the affiliated initial-value problem in the form
\[
\eta(x,t) = \frac{1}{2\pi}\int_{\mathbb{R}}\hat{\eta}_{0}(k) e^{ikx+i\Omega_{+}(k)t} dk + \frac{1}{2\pi}\int_{\mathbb{R}}\hat{\eta}_{0}(k) e^{ikx+i\Omega_{-}(k)t} dk
\]
where
\[
\Omega_{\pm}(k) = \frac{1}{2}\left(\omega s \pm \sqrt{\omega^{2}+4|k|} \right).
\]
If we assume, as per usual when deriving NLS-type models, that 
\[
\hat{\eta}_{0}(k) = \frac{1}{\epsilon}\hat{A}\left(\frac{k-k_{0}}{\epsilon} \right), 
\]
representing the assumption that the initial conditions are narrowly banded around a carrier wave number $k_{0}$, then we see that, taking only the plus branch of $\Omega_{\pm}$, that  
\begin{align*}
\eta(x,t) = & \frac{e^{i\theta(x,t,k_{0})}}{2\pi}\int_{\mathbb{R}}\hat{A}(\tilde{k})e^{i\tilde{k}\xi} e^{i\tau \tilde{\Omega}(k_{0},\tilde{k})}d\tilde{k}\\
\sim & \frac{e^{i\theta(x,t,k_{0})}}{2\pi}\int_{\mathbb{R}}\hat{A}(\tilde{k})e^{i\tilde{k}\xi}e^{i\tau\left(\frac{\Omega''(k_{0})}{2}\tilde{k}^{2} + \frac{\epsilon}{6}\Omega'''(k_{0})\tilde{k}^{3} + \frac{\epsilon^{2}}{24}\Omega''''(k_{0})\tilde{k}^{4}\right)}d\tilde{k},
\end{align*}
where
\[
\theta(x,t,k_{0}) = k_{0}x + \Omega(k_{0})t, ~c_{g}(k_{0}) = \Omega'(k_{0}), ~ \xi = \epsilon(x+c_{g}t), ~ \tau = \epsilon^{2}t, 
\]
and where
\[
\tilde{\Omega}(k_{0},\tilde{k}) = \frac{1}{\epsilon^{2}}\left(\Omega(k_{0}+\epsilon \tilde{k})-\Omega(k_{0})-\Omega'(k_{0})\epsilon\tilde{k}\right) 
\]
and where we have expanded the dispersion relationship up to $\mathcal{O}(\epsilon^{3})$ as is done in the main body of the text.  This expansion corresponds to the affiliated linear-evolution equation for the slowly-evolving envelope $A(\xi,\tau)$
\[
i\pd_{\tau}A = \left(\frac{\Omega''(k_{0})}{2}\pd_{\xi}^{2}-\frac{i\epsilon}{6}\Omega'''(k_{0})\pd_{\xi}^{3} - \frac{\epsilon^{2}}{24}\Omega''''(k_{0})\pd_{\xi}^{4}\right)A. 
\]

As to the issue of why to include the $\epsilon^{2}$ term, we note that the affiliated group velocity of the slowly-evolving envelope is given by 
\begin{align*}
\frac{d}{d\tilde{k}}\tilde{\Omega} = & \frac{1}{\epsilon}\left(\Omega'(k_{0}+\epsilon \tilde{k}) - \Omega'(k_{0})\right)\\
\sim & \Omega''(k_{0})\tilde{k}\left(1 - \frac{3\epsilon c_{g}^{2}}{s}\tilde{k} + 10 \epsilon^{2}c_{g}^{4}\tilde{k}^{2}\right).
\end{align*}
We see that the full dispersion relationship has exactly one root at $\tilde{k}=0$, so that this root is the only point of stationary phase in the corresponding integral representation of the solution $\eta(x,t)$.  However, using our Taylor series expansion, were we to ignore the $\epsilon^{2}$ term, we would introduce a new stationary-phase point at $\tilde{k}_{\ast}$ where
\[
\tilde{k}_{\ast} = \frac{s}{3\epsilon c_{g}^{2}}.
\]
The presence of this stationary-phase point allows for the accumulation of energy at high-frequencies, which without a corresponding band-limiting requirement introduces non-physical effects into the model.  By including the $\epsilon^{2}$ term though, we remove this spurious high-frequency stationary-phase point, thereby avoiding the inclusion of an otherwise unnecessary bandwidth condition on the slowly-evolving envelope.       
\end{document}
